---
title: "Analytics"
---

Here we can report findings from each of our analyses. 

# EDA

# Statistics
```{r}
# Correct the names for Cattaraugus and St. Lawrance in the dataset
tick_lyme_weather <- tick_lyme_weather %>%
  mutate(
    name = case_when(
      name == "Cattaragus" ~ "Cattaraugus",
      name == "St Lawrance" ~ "St. Lawrance",
      TRUE ~ name))

# Verify the changes
unique(tick_lyme_weather$name)

```

```{r}
#load necessary libraries
library(broom)
library(ggplot2)
library(dplyr)
library(knitr)
```

### Table of all relevent data 
```{r}
# Summarize or find the mean for the data for each of the relevant topics
summary_table = tick_lyme_weather %>%
  group_by(name) %>%
  summarize(
    total_cases = sum(lyme_rate, na.rm = TRUE),            
    total_lyme_count = sum(lyme_count, na.rm = TRUE),     
    total_ticks_collected = sum(total_ticks_collected, na.rm = TRUE),  
    mean_prevalence = mean(lyme_rate, na.rm = TRUE),    
    mean_b_burgdorferi = mean(b_burgdorferi_percent, na.rm = TRUE), 
    n_observations = n()) %>%
  mutate(
    prevalence_percent = mean_prevalence * 100,          
    mean_b_burgdorferi_percent = mean_b_burgdorferi * 100)

# Create the table
summary_table %>%
  select(name, total_cases, total_lyme_count, total_ticks_collected, prevalence_percent, mean_b_burgdorferi_percent, n_observations) %>%
  knitr::kable(digits = 2, caption = "Summary of Variables by County")

```

### Relationship between tick density and average temperature per month 
```{r}
density_vs_avg_temp <- lm(tick_population_density ~ jan_19 + feb_19 + mar_19 + apr_19 + may_19 +jun_19 + jul_19 + aug_19 + sep_19 + oct_19 + nov_19 + dec_19 + jan_20 + feb_20 + mar_20 + apr_20 + may_20 + jun_20 + jul_20 + aug_20 + sep_20 + oct_20 + nov_20 + dec_20+ jan_21 + feb_21 + mar_21 + apr_21 + may_21 + jun_21 + jul_21 + aug_21 + sep_21 + oct_21 + nov_21 + dec_21, data = tick_lyme_weather, na.action = na.exclude) #examine the relationship between tick density and average temperature

broom::tidy(density_vs_avg_temp) |> select(term, estimate, p.value) |>knitr::kable(digits =3)
```
This output demonstrates that many of the months show non-significant results, therefore meaning a weak or unclear association of each month with tick density. However some months in 2019, like June (p-value= 0.015) and August (p-value = 0.031), which have significant positive outcomes, and July (p-value < 0.005) had a negative outcome. The intercept was -53.0 with a large of SE of 1,473, and a p-value of 0.971. Therefore  
NEED TO LOOK AT THE REST OF THE VARIABLES

### Relationship between lyme_count and average temperatures per month
```{r}
lyme_vs_avg_temp <- lm(lyme_count ~ jan_19 + feb_19 + mar_19 + apr_19 + may_19 + jun_19 + jul_19 + aug_19 + sep_19 + oct_19 + nov_19 + dec_19 + jan_20 + feb_20 + mar_20 + apr_20 + may_20 + jun_20 + jul_20 + aug_20 + sep_20 + oct_20 + nov_20 + dec_20+ jan_21 + feb_21 + mar_21 + apr_21 + may_21 + jun_21 + jul_21 + aug_21 + sep_21 + oct_21 + nov_21 + dec_21, data = tick_lyme_weather)

broom::tidy(lyme_vs_avg_temp) |> select(term, estimate, p.value) |>knitr::kable(digits =3)
```
Lyme count has similar findings to tick density, with strong positive relationships in August 2019, and negative relationships in June 2019. The intercept was -17,100 with a (SE: 3592) and a p-value < 0.005. 

### Relationship between lyme_count and population variables
```{r}
lyme_rate_vs_population <- lm(lyme_count ~ tick_population_density + total_ticks_collected, data = tick_lyme_weather)

broom::tidy(lyme_rate_vs_population) |> select(term, estimate, p.value) |>knitr::kable(digits =3)
```
This linear regression analyzes the relationship between Lyme rate and tick population density and total ticks collected. Intercept estimate was 121 (SE: 16.4), with a test statistic of 7.39 and p-value<0.005. Therefore the intercept is highly signficant, however the predictors are not. This demonstrates a weak association between Lyme rates and the variables. 

The calculated square mileage estimate was -0.0305 (SE: 0.0159), with a statistic of -1.92 and a p-value<0.005. 
 
### Relationship between lyme_count and area variables
```{r}
lyme_rate_vs_area <- lm(lyme_count ~ shape_area  + shape_leng + pop2020, data = tick_lyme_weather)

broom::tidy(lyme_rate_vs_area) |> select(term, estimate, p.value) |>knitr::kable(digits =3)

summary(tick_lyme_weather$pop2020)
```
 

### Trying the relationship between lyme_count and area variables, including human population density in 2020
```{r}
lyme_rate_vs_area_human <- lm(lyme_count ~ shape_area  + shape_leng  + pop2020, data = tick_lyme_weather)

broom::tidy(lyme_rate_vs_area_human) |> select(term, estimate, p.value) |>knitr::kable(digits =3)
```


### Relationship between b_burgdorferi_percent and population variables
```{r}
b_burgdorferi_percent_vs_population <- lm(b_burgdorferi_percent ~ lyme_count + tick_population_density + total_ticks_collected, data = tick_lyme_weather)

broom::tidy(b_burgdorferi_percent_vs_population) |> select(term, estimate, p.value) |>knitr::kable(digits =3)
```


### Logarithmic examination of lyme rate versus number of ticks collected and tick population density
```{r}
#Convert everything to a logarithmic format
tick_lyme_weather_log <- tick_lyme_weather |>
  mutate(
    log_total_ticks =log1p(total_ticks_collected), 
    log_tick_density = log1p(tick_population_density),
    log_lyme_count = log1p(lyme_count)
  )

# Check for linear relationship with a scatterplot
ggplot(tick_lyme_weather_log, aes(x = log_tick_density, y = log_lyme_count)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Log Tick Density vs Log Lyme Rate",
       x = "Log Tick Density",
       y = "Log Lyme Rate")

#Examination of log-linear regression 
log_model <- lm(log_lyme_count ~ log_total_ticks + log_tick_density + b_burgdorferi_percent, data = tick_lyme_weather_log)
summary(log_model)


# Analyze seasonal weather effects
tick_lyme_weather_seasonal <- tick_lyme_weather_log %>%
  mutate(across(ends_with(c("19", "20", "21")), log1p, .names = "log_{.col}"))

weather_model <- lm(log_lyme_count ~ jan_19 + feb_19 + mar_19 + apr_19 + may_19 + jun_19 + jul_19 + aug_19 + sep_19 + oct_19 + nov_19 + dec_19 + jan_20 + feb_20 + mar_20 + apr_20 + may_20 + jun_20 + jul_20 + aug_20 + sep_20 + oct_20 + nov_20 + dec_20+ jan_21 + feb_21 + mar_21 + apr_21 + may_21 + jun_21 + jul_21 + aug_21 + sep_21 + oct_21 + nov_21 + dec_21, data = tick_lyme_weather_seasonal)

broom::tidy(weather_model) |> select(term, estimate, p.value) |>knitr::kable(digits =3)
```
The logarithmic analysis found that the total number of ticks found and the B. burgdoferi percent had a positive, significant association with increased lyme disease rates. 
<<<<<<< HEAD
=======

Some interesting findings from looking at the log of the weather variables include that months like  March 2019 ( β=−1.002, p<0.001) and October 2020 (β=1.325, p=0.006) showing strong effects. Meanwhile, months like June 2021 (β=0.848, p=0.001) and July 2021 (β=−1.159, p=0.001)) had contrasting impacts, highlighting intra-annual variability in environmental or biological drivers.These months highlight that tick activity is present whenever the temperature is above freezing, is highly variable. 
The low R^2 value demonstrates significant variability. Therefore further analysis with environmental, demographic and behavioral data would be beneficial. 

# Kaleb Analytic Edits:

```{r}
# Boxplot of 'lyme_count' per 'year'

data =
  read_csv("Data Preparation/combined_data_kf.csv")

# Fix mispells of Cattaraugus and St. Lawrance
data <- data %>%
  mutate(
    name = case_when(
      name == "Cattaragus" ~ "Cattaraugus",
      name == "St Lawrance" ~ "St. Lawrance",
      TRUE ~ name))
```

```{r}

histogram_cases = 
  data |>
  ggplot(aes(x = lyme_count)) +
  geom_histogram(aes(y = ..density..), 
                 binwidth = 1, 
                 fill = "blue",
                 alpha = 0.7, 
                 color = "black") +
  labs(
    title = "Distribution of all cases",
    x = "Cases of Lyme Disease",
    y = "Density"
  ) +
  theme_minimal()


# Boxplot of 'lyme_count' per 'year' with mean
boxplots =
  data |>
  ggplot(aes(x = factor(year), y = lyme_count)) +
  geom_boxplot() +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 20,
    size = 3,
    color = "red",
    fill = "red"
  ) +
  labs(
    title = "Distribution of Lyme Disease Cases by Year",
    x = "Year",
    y = "Lyme Disease Cases"
  ) +
  theme_minimal()

histogram_cases
boxplots

# cases are right skewed

data = 
  data |> 
  mutate(log_cases = log10(lyme_count))

histogram_logcases = 
  data |>
  ggplot(aes(x = log_cases)) +
  geom_histogram(aes(y = ..density..), 
                 binwidth = 0.5, 
                 fill = "blue",
                 alpha = 0.7, 
                 color = "black") +
  labs(
    title = "Distribution of all cases",
    x = "Cases of Lyme Disease",
    y = "Density"
  ) +
  theme_minimal()

histogram_logcases

ggsave(
  filename = "histogram_cases.png",
  plot = histogram_cases,          
  path = "Results"
)

ggsave(
  filename = "boxplots_cases_year.png",
  plot = boxplots,          
  path = "Results"
)

ggsave(
  filename = "histogram_log_cases.png", 
  plot = histogram_logcases, 
  path = "Results"
)

#log10 transformation makes outcome a bit better distributed. 
```


```{r}
# Step 2: Histogram and density plot for average temperature

histogram_temp = 
  data |>
  ggplot(aes(x = avg_temp)) +
  geom_histogram(aes(y = ..density..), 
                 binwidth = 1, 
                 fill = "blue",
                 alpha = 0.7, 
                 color = "black") +
  labs(
    title = "Distribution of Average Temperature",
    x = "Average Temperature",
    y = "Density"
  ) +
  theme_minimal()

histogram_temp

# temperature is normally distributed

ggsave(
  filename = "temperature_distribution.png",
  plot = histogram_temp,          
  path = "Results"
)

```


```{r}
# Step 3: Histogram and density plot for 'prev_badtick'

dis_bad_tick_per = 
  data |>
  ggplot(aes(x = b_burgdorferi_percent)) +
  geom_histogram(aes(y = ..density..), binwidth = 0.05, fill = "green", alpha = 0.7, color = "black") +
  geom_density(alpha = 0.3, fill = "purple") +
  labs(
    title = "Distribution of prev_badtick",
    x = "Prevalence of Bad Tick",
    y = "Density"
  ) +
  theme_minimal()

dis_bad_tick_per

ggsave(
  filename = "tickwithbdistribution.png",
  plot = dis_bad_tick_per,          
  path = "Results"
)

# percentage of ticks sampled carrying lyme-causing bacteria is normally distributed

```

```{r}
# Step 4: Line plot of total cases per year
annual_cases <- data %>%
  group_by(year) %>%
  summarize(total_cases = sum(lyme_count, na.rm = TRUE))

cases_per_year = 
  annual_cases |> 
  ggplot(aes(x = year, y = total_cases)) +
  geom_point(color = "red", size = 2) +
  geom_smooth() + 
  labs(
    title = "Total Case Count by Year",
    x = "Year",
    y = "Total Case Count"
  ) +
  theme_minimal()

cases_per_year
ggsave(
  filename = "cases_per_year.png",
  plot = cases_per_year,          
  path = "Results"
)


```

```{r}
# Regression model of case_count on prev_badtick controlling for avg_temperature
model <- lm(log_cases ~ b_burgdorferi_percent + avg_temp, data = data)
summary(model)

regression_table = 
  tbl_regression(
  model,
  intercept = TRUE,
  label = list(
    b_burgdorferi_percent ~ "% Ticks Carry",
    avg_temp ~ "Avg Temp"
  )
)

regression_table

gt_table = 
  as_gt(regression_table)

gtsave(gt_table, filename = "Results/regress.html")
```

```{r}
# Step 6: Create a summary statistics table (Table 1) for each county
table1 = 
  data |> 
  select(name, lyme_count, avg_temp, b_burgdorferi_percent) |> 
  tbl_summary(
    by = name,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} / {N} ({p}%)"
    ),
    digits = all_continuous() ~ 2
  ) %>%
  add_overall() %>%
  bold_labels() 

# Display the summary table
table1

library(dplyr)
library(gt)

# Calculate summary statistics for each county
summary_stats <- data %>%
  group_by(name) %>%
  summarize(
    years_with_data = n(),
    lyme_count_mean = mean(lyme_count, na.rm = TRUE),
    avg_temp_mean = mean(avg_temp, na.rm = TRUE),
    avg_temp_sd = sd(avg_temp, na.rm = TRUE),
    b_burgdorferi_percent_mean = mean(b_burgdorferi_percent, na.rm = TRUE),
    b_burgdorferi_percent_sd = sd(b_burgdorferi_percent, na.rm = TRUE)
  ) %>%
  ungroup()

# Create a gt table
summary_table <- summary_stats %>%
  gt(rowname_col = "name") %>%
  fmt_number(
    columns = vars(years_with_data, lyme_count_mean, avg_temp_mean, avg_temp_sd, b_burgdorferi_percent_mean, b_burgdorferi_percent_sd),
    decimals = 2
  ) %>%
  cols_label(
    years_with_data = "Non-Missing Years",
    lyme_count_mean = "All Year Mean Lyme Cases",
    avg_temp_mean = "Mean Temperature",
    avg_temp_sd = "SD",
    b_burgdorferi_percent_mean = "Mean B. burgdorferi % ",
    b_burgdorferi_percent_sd = "SD"
  ) %>%
  tab_header(
    title = "Summary Statistics by County"
  )

# Display the table
summary_table

gtsave(data = summary_table, filename = "Results/table.1.html")

```
